Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2017-04-23 article/article_v0.9.5.tex
First complete version of the article is done.

\documentclass[titlepage,11pt,twoside]{article}

\usepackage[dvips]{graphicx}

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{multirow}
%\usepackage{colortbl}
%\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}

%\usepackage{submit}

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
%\newcommand{\argmax}{\text{argmax}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

@@

\linespacing{1}

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}

\author{Anne H. Petersen, Bo Markussen, Karl Bang Christensen}

\affil{University of Copenhagen}

@@

\linespacing{1}

@@

\begin{abstract}
abstract...
\begin{keywords}
multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), goodness of fit, European Social Survey.
\end{keywords}
\end{abstract}

@@

Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well. However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt.\ data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable? Three examples of this are

\begin{itemize}
\item Surveys often use mixed modes of administration, e.g. mail and telephone, and while this can improve response rates, the mode of administration can affect results \citep{Brambilla1987,McHorney1994}, and differences in response behaviour can lead to biased results. Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
\item The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population. However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions), see \cite{Liu2016}.
\item Large scale open source data sets such as the PISA data and ESS (European Social Survey) data. In these data sets, data analysts are far away from the data producers and problem-specific recommendations about potential instrument-induced challenges in the data sets are not available for the data analysts.
\end{itemize}

\hl{Maybe broaden the raison d'etre of PCADSC a bit: It is also a useful tool for doing initial, exploratory data analysis and it helps us to identify potential problems or challenges early in the study design development, before concrete model choices have been made, thereby aiding a more economic use of resources and a less ad-hoc approach to the types of problems that PCADSC can find.}

Assume that we have two data sets with the same variables, but different observations represented as a data set with a subset-inducing variable. Assume next that we wish to compare structures without specifying a model, or even a variable of interest. The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be taken into account.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model \hl{Skal der referencer her? Nogen forslag?}, but when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables. The use of parametric models does not constitute general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and not to the interplay between variables. Entry-by-entry comparison of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase. Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.

%For these methods significant $p$-values are likely to abound as the sample size increases. Bo: Jeg tror denne sætning bare kan udelades.

We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. This yields a solution that is largely independent of the sizes of the two subsets of data. The proposed tools are implemented in the statistical software \texttt{R} (\cite{R}) statistical software in our package, \texttt{PCADSC}, which is available at \hl{? How to refer to a package on Github?}.

This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA) in general and presents a worked data example using open source, online available data on psychological well-being in three European countries. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries.

\section{PCA-based tools for data structure comparisons}
\label{sec:pcadscintro}
As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a data set. But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context. One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. This gives it a very nice interpretation as a measure of the overall structure. If we do not accept the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.

A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry. Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once. However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance. In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data. On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.

At this point a general remark about standardization is in place. PCA deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables often are standardized prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation. In this paper we will adopt this approach, and in particular, we will allow for differences between scalings of the variables in the two data subsets.

\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?}

\subsection{A brief introduction to Principal component analysis}
Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$. We assume in the following that all variables of $X$ have a numerical interpretation, e.g. by being continuous or ordinal and categorical. For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost. More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one. Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.

More specifically, the rank-q-reconstruction error is attained when we choose
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$. In the PCA framework, we refer to the eigenvectors as \textit{loadings}. The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
\begin{equation*}
\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
\end{equation*}
where $X_j \in \RR^n$ \hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function. This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe. And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference. If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.

The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
where the initial subspace is defined as $\hat{K}_0 = \{0\}$. It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and the variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension, $q$, a priori.

\subsection{Using PCA for data structure comparisons}
All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful quantities to compare across such different datasets. Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings. But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets. This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variables are used differently in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. For each of these two datasets, we complete the following steps:
\begin{enumerate}
\item Standardized each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets.
\item Form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.}
\end{equation*}
thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1,\dotsc,d$.
\end{enumerate}
The hereby obtains PCA decompositions of the correlation matrices can then be compared. We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

The standardization makes the variables comparable on the same scale (units of standard deviation), and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
\begin{equation*}
\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
\end{equation*}
This identity will simplify some expressions below.

@@

The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in (\cite{ESStopline5}) and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

>>>>

\documentclass[titlepage,11pt,twoside]{article}
\usepackage[dvips]{graphicx}

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
%\usepackage{pmbib}

\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{multirow}
%\usepackage{colortbl}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage[T1]{fontenc}

\usepackage{caption,setspace}
\captionsetup{font={footnotesize,stretch=1.5}}
\renewcommand{\baselinestretch}{2}

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

@@

\begin{titlepage}
\linespacing{1}

\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}

\author{Anne H. Petersen}
\affil{Department of Public Health, University of Copenhagen}

\author{Bo Markussen}
\affil{Department of Mathematical Sciences, University of Copenhagen}

\author{Karl Bang Christensen}
\affil{Department of Public Health, University of Copenhagen}

@@

\end{titlepage}\vspace*{24pt}

@@

\begin{abstract}
\linespacing{1}

In empirical psychometric work, one often encounters data that is somehow divided into two distinct susbets. This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc. Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are fundamentally similar in structures, that is, if they measure the same underlying concepts. However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc and require making full model assumptions. We present a suite of visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC). This approach requires no directional nor hierarchical assumptions on the data. Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets and by each targeting a different aspect of the PCA information, the three plots presented here give a thorough and deep understanding of where differences occur and why it might be the case.

The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria. Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g. mean scores in different well-being aspects is not meaningful. However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being. Therefore, inter-country comparisons are warranted for these two countries.

\begin{keywords}
multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
\end{keywords}
\end{abstract}

@@

Data comparability is a classical topic in psychometrics and, more generally, in statistics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assesed empirically. This happens e.g. when data are collected across several centers (or even countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied. All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of collapsibility among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment very ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a valid general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables. On the other hand, e.g. entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.

While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection. Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well. However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies. This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for all to use. However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g. potential instrument-induced challenges in the datasets are not available for these data analysts. For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.

The data structure comparison problem is however not limited to the context of new data sharing platforms. Even in single-population survey studies, a mixture of modes of administration is often used, e.g. mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results. Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle. Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e. pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).

The combarability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable. Assume next that we wish to compare data structures without specifying a model, or even a variable of interest. The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.

We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. 
%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.

This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g. when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project.

\section{PCA-based tools for data structure comparisons}
\label{sec:pcadscintro}
As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset. But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context. One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. This gives it a very nice interpretation as a measure of the overall structure. If we do not accept the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.

A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry. Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once. However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance. In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data. One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.

At this point a general remark about standardization is in place. PCA deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data. Therefore, the variables are typically standardized prior to performing PCA. Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices. The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation. In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.

%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.

\subsection{A brief introduction to principal component analysis}
Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$. We assume in the following that all variables of $X$ have a numerical interpretation, e.g. by being continuous or ordinal and categorical. For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost. More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one. Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}. More specifically, the rank-q-reconstruction error is attained when we choose
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$. In the PCA framework, we refer to the eigenvectors as \textit{loadings}. The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
\begin{equation*}
\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
\end{equation*}
where $x_j \in \RR^n$ 
%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function. This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe. And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference. If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.

The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
where the initial subspace is defined as $\hat{K}_0 = \{0\}$. It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

\subsection{Using PCA for data structure comparisons}
All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings. But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets. This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. For each of these two datasets, we complete the following steps:
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets.
\item Form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.}
\end{equation*}
thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
\begin{equation*}
\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
\end{equation*}
This identity will simplify some expressions below.

@@

The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
